---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: Aidan Power
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#Check working directory
getwd()

#Load packages
library(tidyverse)
library(agricolae)
library(lubridate)
library(here)

#Import Raw File
ChemPhys <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                     stringsAsFactors = TRUE)

#Convert date from factor
ChemPhys$sampledate <- as.Date(ChemPhys$sampledate, format = "%m/%d/%y")

#2
#Set theme
mytheme <- theme_classic(base_size = 10) +
  theme(axis.text = element_text(color = "blue"), 
        legend.position = "right")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature does not change with depth across all lakes in July. 
Ha: Mean lake temperature does change with depth across all lakes in July.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

ChemPhys_Updated <- ChemPhys %>%
  mutate(Month = month(ymd(sampledate))) %>% 
  filter(Month == 7) %>% #Filter for July
  select(lakename, year4, daynum, depth, temperature_C) %>% #Selecting certain columns
  drop_na() #Dropped NAs


#5
tempbydepth <- 
  ggplot(ChemPhys_Updated, aes(x = depth, y = temperature_C)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ylim(0, 35) +
  labs(title = "Temperature vs. Depth in July",x="Depth(m)",y="Temperature(C)") 

print(tempbydepth)


  
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
As depth increases, it appears that temperature decreases. Although a linear model works, I would recommend an exponential one, as it appears to fit the shape better. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
tempdepth.regression <- 
  lm(ChemPhys_Updated$temperature_C ~ 
       ChemPhys_Updated$depth)
summary(tempdepth.regression) 

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The variability in temperature is mostly explained by change in depth, as R-squared is 0.7387. 9726 degrees of freedom means that there was a sample size of 9727, which is large. The p value is also much smaller than 0.05, which means that temperature and depth have a very significant negative relationship. For every meter, temperature is expected to decrease.  

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
ChemPhysAIC <- lm(data = ChemPhys_Updated, temperature_C ~ year4 + daynum + depth)

#Choose a model by AIC in a Stepwise Algorithm
step(ChemPhysAIC)


#10
ChemPhysModel <- lm(data = ChemPhys_Updated, temperature_C ~ year4 + daynum + depth)
summary(ChemPhysModel)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The final set of explanatory variables that the AIC method suggests we use is the same as above (year4, daynum, depth). This model appears to explain 74.11% of the variance, and compared to the other model, which explained only 73.87%, the multiple regression is an improvement (albeit not by much). 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#ANOVA models
ChemPhys.anova <- aov(data = ChemPhys_Updated, temperature_C ~ lakename)
summary(ChemPhys.anova)

#Linear Model
ChemPhys.anova2 <- lm(data = ChemPhys_Updated, temperature_C ~ lakename)
summary(ChemPhys.anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
There is significant difference in mean temperature among the lakes, as noted by low p-values for both the ANOVA and linear model, as well as each lake having its own significant p-value. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
tempbydepth.lake <- 
  ggplot(ChemPhys_Updated,aes(x=depth,y=temperature_C,color=lakename)) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0,35) +
  labs(title="Temperature by Depth in July Sorted by Lake",
       x="Depth(m)",y="Temperature(C)",color="Lake Name") +
  theme(legend.position = "right")
  
print(tempbydepth.lake)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(ChemPhys.anova)

#Extracting groupings to view relationships
ChemPhys.Totals.groups <- HSD.test(ChemPhys.anova, "lakename", group = TRUE)
print(ChemPhys.Totals.groups)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
From the findings above, Ward Lake and Paul Lake have the same mean temperature. No lake has a mean temperature that is statistically distinct from all other lakes. 
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 
If we were just looking at Peter and Paul Lake, we could use a two-sample t-test to see whether they have distinct mean temperatures. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#Filter for the two lakes
ChemPhys_CrampWard <- filter(ChemPhys_Updated,lakename %in% c("Crampton Lake","Ward Lake"))

#t-test
CrampWard.twosample <- t.test(ChemPhys_CrampWard$temperature_C ~ ChemPhys_CrampWard$lakename)
print(CrampWard.twosample)
```

>Answer: 
The test says that the mean temperatures for these two lakes are not significantly different from one another. Therefore, the mean temperatures for the lakes are essentially equal. This matches the idea that no lake is statistically significant from all other lakes, as Crampton and Ward are not statistically significant from each other (0.2649>0.05). 

